---
title: "Plotting CMIP6 ESM data in R"
author: "Denisse Fierro Arcos"
date: "2022-10-29"
output: 
  github_document:
    toc: true
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

In this notebook, we will use the `metR` and `ggplot2` libraries in `R` to load the monthly means from the CMIP6 data.


## Loading libraries

```{r libraries, results = "hide", warnings = F, message = F}
library(metR)
library(tidyverse)
```

## Loading search results from [first notebook](01_Querying_CMIP6_database.md)
We will use the search results from the first notebook to find the files we need. We will only focus on the `intpp` variable and the `ssp245` scenario.

```{r}
#Loading search results
results_query <- read_csv("../Outputs/results_merged.csv") %>%
  filter(variable_id == "intpp" & experiment_id == "ssp245") %>% 
  #We will remove any duplicated entries - based on the output file name
  mutate(dup = duplicated(out_file_name)) %>%
  filter(dup == F) %>% 
  #Add name of output files of interest
  mutate(mean_dec0 = list.files(out_full_folder, pattern = paste0("*.intpp.*Mean.", decade_0_start), full.names = T),
         mean_decN = list.files(out_full_folder, pattern = paste0("*.intpp.*Mean.", decade_N_start), full.names = T))

head(results_query, n = 2)
```



## Loading data

```{r}
#Loading raster with means from first decade
raster_stack <- function(file_dec0, file_decN, var_id, months_ind){
  ReadNetCDF(file_dec0, vars = var_id) %>% 
  #Extracting months of interest
  filter(month %in% months_ind) %>% 
  #Adding a column indicating to which decade value relates to
  mutate(decade = "Decade_0") %>% 
  #Loading raster with means from last decade
  bind_rows(ReadNetCDF(file_decN, vars = var_id) %>% 
              #Extracting months of interest
              filter(month %in% months_ind) %>% 
               #Adding a column indicating to which decade value relates to
              mutate(decade = "Decade_N")) %>%
  #Calculating difference between first and last decade
  pivot_wider(names_from = decade, values_from = var_id) %>%
  mutate(Difference = Decade_0 - Decade_N,
         x = case_when(x >180 ~ x-360,
                       T~x)) %>%
  group_by(month)
  }

#Function creating one raster per month of interest - It will work with one or multiple decades
raster_month <- function(df){
  ncols_df <- ncol(df)
  df <- df %>% 
    ungroup() %>% 
    dplyr::select(-month) %>% 
    #Changing data frame to sf object
    st_as_sf(coords = c("x", "y"), crs = 4326) %>% 
    #Creating raster
    st_rasterize() 
  
  if(ncols_df > 4){
    df_final <- df %>% 
    #Changing attributes to dimensions
    st_redimension() %>% 
    #Renaming dimension to decades. Value 2 represents difference between first (0) and last (1) decade
    st_set_dimensions("new_dim", values = c(0,1,2), names = "decade")
  }else{
      df_final <- df
    }
  return(df_final)
  }
```


```{r}
months_ind <- c(1, 9)

test <- results_query %>% 
  dplyr::select(variable_id, mean_dec0, mean_decN, source_id) %>% 
  as_tibble() %>% 
  mutate(df = NA, raster_dec0 = NA, raster_decN = NA, raster_diff = NA, raster_all = NA, figure = NA)

#Facet labels
label_names <- c(`0` = "First decade",
                   `1` = "Last decade",
                   `2` = "Difference")

for(i in 1:nrow(test)){
  #Save data frame with combined datasets
  test$df[i] <- list(raster_stack(test$mean_dec0[i], 
                                  test$mean_decN[i], 
                                  test$variable_id[i], 
                                  months_ind))
  #Save one raster per month/decade using data frames from previous step
  ras <- test$df[i][[1]] %>% 
    #Make dataset longer and group by decade
    ungroup() %>% 
    pivot_longer(Decade_0:Difference, names_to = "Decade") %>% 
    group_by(Decade, month) %>% 
    #Save intermediate output to get names for list items
    {. ->> temp} %>% 
    #Split into groups based on decades
    group_split() %>% 
    map(dplyr::select, -Decade) %>% 
    #Save as rasters
    map(raster_month) %>% 
    #Rename each list item before allocating to tibble columns
    setNames(group_keys(temp) %>% unite("key", Decade:month) %>% pull())
  #Allocating rasters to correct column
  raster_dec0[i] <- list(Jan = ras$Decade_0_1,
                         Sep = ras$Decade_0_9)
  raster_decN[i] <- list(Jan = ras$Decade_N_1,
                         Sep = ras$Decade_N_9)
  raster_diff[i] <- list(Jan = ras$Difference_1,
                         Sep = ras$Difference_9)
  
  #Save one raster with all data merged together
  test$raster_all[i] <- list(group_split(test$df[i][[1]]) %>%
                               map(raster_month) %>% 
                               setNames(month.abb[group_keys(test$df[1][[1]]) %>% pull()]))
  #Saving figures per decade
  figs <- list(jan_dec0 = ggplot()+
                 geom_stars(data = raster_dec0[i][[1]]$Jan)+
                 scale_fill_distiller(palette = "BuGn", direction = 1,
                                     guide = guide_colorbar(title = test$variable_id[i],
                                             title.position = "top",
                                             title.hjust = 0.5))+
                 coord_sf()+
                 labs(title = test$source_id[i])+
                 theme_bw()+
                 theme(legend.position = "bottom",
                       legend.key.width = unit(3, "cm"),
                       axis.title = element_blank(), 
                       panel.ontop = T, 
                       panel.background = element_rect(color = NA, fill = NA),
                       panel.grid = element_line(color = "orange", size = 0.3, linetype = "dotted")
                       ),
    jan_decN = jan_fig = ggplot()+
                 geom_stars(data = raster_decN[i][[1]]$Jan)+
                 scale_fill_distiller(palette = "BuGn", direction = 1,
                                     guide = guide_colorbar(title = test$variable_id[i],
                                             title.position = "top",
                                             title.hjust = 0.5))+
                 coord_sf()+
                 labs(title = test$source_id[i])+
                 theme_bw()+
                 theme(legend.position = "bottom",
                       legend.key.width = unit(3, "cm"),
                       axis.title = element_blank(), 
                       panel.ontop = T, 
                       panel.background = element_rect(color = NA, fill = NA),
                       panel.grid = element_line(color = "orange", size = 0.3, linetype = "dotted")))
  test$figureJan[i] <- list(figs)
}

```

