---
title: "Accessing and downloading CMIP6 data"
author: "Denisse Fierro Arcos"
date: "2022-10-28"
output: 
  github_document:
    toc: true
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
Based on the results query obtained from our [previous notebook](01_Querying_CMIP6_database.md), we will download the raw CMIP6 files using `R`. Then, we will switch to `Python` to make use of the `xmip` library that allow us to standardise variable names for CMIP6 data.

## Accessing and downloading CMIP6 data
First, we will load the `R` libraries we will need for this notebook.

```{r}
library(reticulate)
library(tidyverse)
```

We will now load the query results we obtained from the previous notebook. Remember that we added a number of columns that could help us select a subset of files that are relevant to us. For example, if we want only files that linked to the first decade of the dataset, we could use the `decade` column, where `dec0` means that a particular file contains data for the first decade, and `both` means that the file has the first and last decades. 

```{r}
#Loading the results
results_query <- read_csv("../Outputs/results_merged.csv")

#Activate if needed - Ignores any files outside the first and last decades of the dataset
# results_query <- results_query %>% 
#   filter(keep == T)
```

We will now download all files into our disk. We will put them in a temporary folder, which we will remove after we have process all files.

```{r}
for(i in 1:nrow(results_query)){
  #Creating the temporary folder where raw files will be saved
  path_out <- file.path(results_query$out_full_folder[i], "raw")
  #Using their original file name
  fn <- str_extract(results_query$file_id[i], ".*[0-9].nc")
  #If temporary file does not exist, create it
  if(file.exists(path_out) == F){
    dir.create(path_out, recursive = T)
  }
  #Downloading file and saving into temporary folder
  curl::curl_download(results_query$file_url[i], file.path(path_out, fn))
  }

```

# Switching to `Python`
The `reticulate` package in `R` allow us to use `Python` within an `R` script. We will be using the `CMIP6_data` conda environment, which in simple terms is a directory that contains all the `Pyhton` libraries and the dependencies needed to run this script. The `README` file in this repository has instructions on how to create this environment using the `yml` file provided in this repository.

```{r , results = "hide", warnings = F, message = F}
#Activating the conda environment containing relevant Python libraries
use_condaenv("CMIP6_data")
```

## Loading `Python` libraries
```{python}
#Loading and manipulating netcdf files
import xarray as xr
import numpy as np
import pandas as pd

#Standardisation of CMIP6 data for easy data post-processing
from xmip.preprocessing import rename_cmip6, promote_empty_dims, broadcast_lonlat, correct_coordinates

#Dealing with file paths
import os
from glob import glob
import re

#Plotting
import matplotlib.pyplot as plt
```

### Transforming CMIP6 search results into a `Python` variable
In this step we can use either of the search results. In the example below, we will be working with the merged search results.
```{python}
#Load data and ensure there are no duplicates
CMIP6_query = pd.read_csv("../Outputs/results_merged.csv").drop_duplicates()

CMIP6_query = CMIP6_query.astype({'decade_0_start': 'int32', 'decade_0_end': 'int32',\
'decade_N_start': 'int32', 'decade_N_end': 'int32'})

CMIP6_query.head(n = 2)
```

### Accessing and saving CMIP6 data
First, we will define a new function that will download the CMIP6 data of interest.

```{python}
def load_CMIP6(url, var_id):
  #Loading data to memory in a temporary file
  link = requests.get(url).content
  remote = Dataset(var_id, memory = link)
  #Loading as xarray dataset
  sub = xr.open_dataset(xr.backends.NetCDF4DataStore(remote))
  return sub

```

We will define a second function that will standardise all variable names in CMIP6 model outputs before saving a copy locally.

```{python}
#Defining function to standardise CMIP6 data
def standard(dataset):
  #Renaming all variables so they are the same for all models
  dataset = rename_cmip6(dataset)
  #Adding an index if not included
  dataset = promote_empty_dims(dataset)
  #Converting coordinates from 1D to 2D arrays
  dataset = broadcast_lonlat(dataset)
  #Correcting coordinates if needed so they appear in format 0-360 degrees
  dataset = correct_coordinates(dataset)
  return dataset

```

We will now define a function that will access the CMIP6 data, load into memory, extract years of interest, and standardise all variable names before saving a copy locally.
```{python}
def download_CMIP6(df):
  #Getting key information to load CMIP6 data locally from query results
  #URL address (include "#mode=bytes" at the end of the URL if not working)
  url = df['file_url']
  #Variable name
  var_id = np.unique(df['variable_id']).tolist()
  #Start decades
  dec_0 = int(np.unique(df['decade_0_start']))
  dec_N = int(np.unique(df['decade_N_start']))
  #File paths
  out_folder = np.unique(df['out_full_folder']).tolist()
  #Ensuring folder exists
  os.makedirs(out_folder[0], exist_ok = True)
  #File name with full path
  out_file = np.unique(df['out_path']).tolist()
  
  # ds = []
  # for ind in df.index:
  #   sub = load_CMIP6(url[ind], var_id)
  #   ds.append(sub)
  # del sub
  # ds = xr.concat(ds, dim = 'time').chunk({"time": 10})
  # 
  # if len(df) == 1:
  #   #Subsetting data - First and last decade and stitching them together
  #   d0 = ds.sel(time = slice(str(dec_0), str(dec_0+9)))
  #   dN = ds.sel(time = slice(str(dec_N), str(dec_N+9)))
  #   #Combining first and last decade into one dataset
  #   ds_sub = xr.concat([d0, dN], dim = 'time')
  # elif df.decade.unique() == 'dec0':
  #   ds_sub = ds.sel(time = slice(str(dec_0), str(dec_0+9)))
  # else:
  #   ds_sub = ds.sel(time = slice(str(dec_N), str(dec_N+9)))
  # del ds
  
  for ind in df.index:
    ds = load_CMIP6(url[ind], var_id)
    if len(df) == 1:
      #Subsetting data - First and last decade and stitching them together
      d0 = ds.sel(time = slice(str(dec_0), str(dec_0+9)))
      dN = ds.sel(time = slice(str(dec_N), str(dec_N+9)))
      #Combining first and last decade into one dataset
      ds_sub = xr.concat([d0, dN], dim = 'time')
      del d0, dN, ds
      #Applying standardisation
      ds_sub = standard(ds_sub)
      #Saving subsetted dataset
      ds_sub.to_netcdf(out_file[0])
    else:
      fn = os.path.join(out_folder[0], re.split("\\|", df.file_id[ind])[0])
      #Applying standardisation
      ds = standard(ds)
      #Saving subsetted dataset
      ds.to_netcdf(fn)

```

We will loop through each item in the search results.
```{python, eval = F}
#Starting loop through each model
for mod in np.unique(CMIP6_query['source_id']):
  print(mod)
  mod_id = CMIP6_query[CMIP6_query['source_id'] == mod]
  #Group by variable name, experiment ID and grid type before starting data download
  for names, sub in mod_id.groupby(['dataset_id', 'decade']):
    print(sub[['experiment_id', 'grid_label', 'variable_id']])
    download_CMIP6(sub.reset_index(drop = True))

```

### Checking subsetted data
We can load one of the datasets we saved in the previous step to check its contents.

```{python}
#Loading last dataset saved locally
ds = xr.open_dataset(CMIP6_query['out_path'][47])
#Checking contents
ds
```

### Plotting results
Finally, we can calculate monthly means over the first decade of interest and plot these results.
```{python}
#Selecting the first decade
ds = ds.sel(time = str(CMIP6_query['decade_0_start'][47]))
#Calculating monthly means and plotting only the first month
ds[CMIP6_query['variable_id'][47]].mean('time').plot(levels = 9)
#Show plot
plt.show()
#Close plot if needed
#plt.close()
```

In this notebook, we have successfully downloading a subset of the CMIP6 datasets that we identified in [the first notebook](01_Querying_CMIP6_database.md). Next, we will describe how to calculate monthly means from the datasets that we saved locally.

