---
title: "Accessing and downloading CMIP6 data"
author: "Denisse Fierro Arcos"
date: "2022-10-28"
output: 
  github_document:
    toc: true
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(reticulate)
```


## Accessing and downloading subsets of CMIP6 data
We will now switch to `Python` to make use of the `xmip` library that allow us to standardise variable names for CMIP6 data.

### Switching to `Python`
The `reticulate` package in `R` allow us to use `Python` within an `R` script. We will be using a `conda environment` which in simple terms is a directory that contains all the `Pyhton` libraries and the dependencies needed to run this script. The `README` file in this repository has instructions on how to create this environment.

```{r , results = "hide", warnings = F, message = F}
#Activating the conda environment containing relevant Python libraries
use_condaenv("CMIP6_data")
```

### Loading `Python` libraries
```{python}
#Easy access to data hosted online
import requests

#Loading and manipulating netcdf files
from netCDF4 import Dataset
import xarray as xr
import numpy as np
import pandas as pd

#Standardisation of CMIP6 data for easy data post-processing
from xmip.preprocessing import rename_cmip6, promote_empty_dims, broadcast_lonlat, correct_coordinates

#Dealing with file paths
import os
from glob import glob

#Plotting
import matplotlib.pyplot as plt
```

### Transforming CMIP6 search results into a `Python` variable
In this step we can use either of the search results. In the example below, we will be working with the merged search results.
```{python}
#Load data and ensure there are no duplicates
CMIP6_query = pd.read_csv("../Outputs/results_merged.csv").drop_duplicates()

CMIP6_query = CMIP6_query.astype({'decade_0_start': 'int32', 'decade_0_end': 'int32',\
'decade_N_start': 'int32', 'decade_N_end': 'int32'})

CMIP6_query.head(n = 2)
```

### Accessing and saving CMIP6 data
We will define a new function that will access the CMIP6 data, load into memory, extract years of interest and standardise all variable names before saving a copy locally.

First, we will define a function that will standardise all variable names in CMIP6 model outputs before saving a copy locally.

```{python}
#Defining function to standardise CMIP6 data
def standard(dataset):
  #Renaming all variables so they are the same for all models
  dataset = rename_cmip6(dataset)
  #Adding an index if not included
  dataset = promote_empty_dims(dataset)
  #Converting coordinates from 1D to 2D arrays
  dataset = broadcast_lonlat(dataset)
  #Correcting coordinates if needed so they appear in format 0-360 degrees
  dataset = correct_coordinates(dataset)
  return dataset

```

Now we will define the download function, which will include our previously defined standardisation function.
```{python}
def download_CMIP6(df):
  #Getting key information to load CMIP6 data locally from query results
  #URL address (include "#mode=bytes" at the end of the URL if not working)
  url = np.unique(df['file_url'])
  #Variable name
  var_id = np.unique(df['variable_id']).tolist()
  #Start decades
  dec_0 = int(np.unique(df['decade_0_start']))
  dec_N = int(np.unique(df['decade_N_start']))
  #File paths
  out_folder = np.unique(df['out_full_folder']).tolist()
  #Ensuring folder exists
  os.makedirs(out_folder[0], exist_ok = True)
  #File name with full path
  out_file = np.unique(df['out_path']).tolist()
  
  ds = []
  for ind in df.index:
    #Loading data to memory in a temporary file
    link = requests.get(url[ind]).content
    remote = Dataset(var_id, memory = link)
    #Loading as xarray dataset
    sub = xr.open_dataset(xr.backends.NetCDF4DataStore(remote))
    ds.append(sub)
  ds = xr.concat(ds, dim = 'time')
    
  #Standardising CMIP6 data
  ds = standard(ds)
  
  #Subsetting data - First and last decade and stitching them together
  d0 = ds.sel(time = slice(str(dec_0), str(dec_0+9)))
  dN = ds.sel(time = slice(str(dec_N), str(dec_N+9)))
  
  #Combining first and last decade into one dataset
  ds_sub = xr.concat([d0, dN], dim = 'time')
  
  #Saving subsetted dataset
  ds_sub.to_netcdf(out_file[0])

```

We will loop through each item in the search results.
```{python, eval = F}
#Starting loop through each model
for mod in np.unique(CMIP6_query['source_id']):
  print(mod)
  mod_id = CMIP6_query[CMIP6_query['source_id'] == mod].reset_index()
  #Group by variable name, experiment ID and grid type before starting data download
  for names, sub in mod_id.groupby(['experiment_id', 'grid_label', 'variable_id']):
    print(sub[['experiment_id', 'grid_label', 'variable_id']])
    download_CMIP6(sub.reset_index(drop = True))

```

### Checking subsetted data
We can load one of the datasets we saved in the previous step to check its contents.

```{python}
#Loading last dataset saved locally
ds = xr.open_dataset(CMIP6_query['out_path'][1])
#Checking contents
ds
```

### Plotting results
Finally, we can calculate monthly means over the first decade of interest and plot these results.
```{python}
#Selecting the first decade
ds = ds.sel(time = str(CMIP6_query['decade_0_start'][1]))
#Calculating monthly means and plotting only the first month
ds[CMIP6_query['variable_id'][1]].mean('time').plot(levels = 9)
#Show plot
plt.show()
#Close plot if needed
#plt.close()
```

In this notebook, we have successfully downloading a subset of the CMIP6 datasets that we identified in [the first notebook](01_Querying_CMIP6_database.md). Next, we will describe how to calculate monthly means from the datasets that we saved locally.

