---
title: "Calculating monthly means in Python using CMIP6 ESM data"
author: "Denisse Fierro Arcos"
date: "2022-10-25"
output: 
  github_document:
    toc: true
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
In this notebook, we will use the `xarray` library in `Python` to calculate monthly means using the CMIP6 data downloaded in the previous step: [01_Accessing_CMIP6_data](01_Accessing_CMIP6_data.md). We will then save the results as a netcdf files locally.


```{r libraries, results = "hide", warnings = F, message = F}
library(reticulate)
use_condaenv("CMIP6_data")
```

# Loading search results from previous step
In the first notebook, we added a new column that points to the location where the CMIP6 data we are interested in where saved. We will load them to our environment before we calculate the monthly means for each variable.

We will 

```{python}
import xarray as xr
import os
from glob import glob
```

```{python}
base_dir = "/perm_storage/home/data/CMIP6_data/*/*/*.nc"
x = glob(base_dir)
y = [a for a in x if "MonthlyMean" not in a]

test = xr.open_dataset(y[0])
test.intpp.sel(time = slice('2015', '2024')).groupby('time.month').mean('time').plot(col = 'month', )
matplotlib.pyplot.show()
```


## Calculating monthly means per decade
For this portion of the script, we will switch back to `R`. We will calculate monthly means per decade and save them locally.

