---
title: "Calculating monthly means in R using CMIP6 ESM data"
author: "Denisse Fierro Arcos"
date: "2022-10-24"
output: 
  github_document:
    toc: true
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
In this notebook, we will use the `ncdf4` library in `R` to calculate monthly means using the CMIP6 data downloaded in the previous step: [02_Downloading_CMIP6_data](02_Downloading_CMIP6_data.md). We will then save the results as a netcdf files locally.


```{r libraries, results = "hide", warnings = F, message = F}
library(tidyverse)
library(lubridate)
library(ncdf4)
```

## Loading search results from [first notebook](01_Querying_CMIP6_database.md)
In the first notebook, we added a new column that points to the location where the CMIP6 data we are interested in where saved. We will load them to our environment before we calculate the monthly means for each variable.

Recall from the previous step, some models had multiple files available for the `intpp` variable. We will remove any duplicate entries as we saved a single file with the initial and last decade per model.

```{r}
#Loading ACCESS-ESM 1.5 query
results_query <- read_csv("../Outputs/results_merged.csv") %>% 
  #We will remove any duplicated entries - based on the output file name
  mutate(dup = duplicated(out_file_name)) %>%
  filter(dup == F)

head(results_query, n = 2)
```


## Calculating monthly means per decade
There are several `R` packages that could be used to calculate monthly means from `netcdf` files with ease, but most do not offer the option of saving the results as `netcdf` files.
  
Here we use the `ncdf4` package to achieve this. The resulting workflow is rather complicated, so if you prefer a more streamlined option, refer to the [`Python` version](02b_Calculating_monthly_means_in_Python.Rmd) of this notebook.

```{r, eval = F}
# We will loop through the results data frame
for(i in 1:nrow(results_query)){
  #Loading raster
  r <- nc_open(results_query$out_path[i])
  #Accessing variable of interest
  var_id <- results_query$variable_id[i]
  var <- ncvar_get(r, var_id)
  #Accessing coordinates - notice these are 2D arrays
  lon <- ncvar_get(r, "lon")
  lat <- ncvar_get(r, "lat")
  #Creating 1D index for coordinates for easy plotting
  #longitude coordinates formatted -180 to +180
  x <- ifelse(lon[,1,1] > 180, lon[,1,1]-360, lon[,1,1])
  y <- lat[1,,1]
  #Accessing time dimension
  time <- ncvar_get(r, "time")
  #Reordering variables along x axis for easy plotting
  var <- var[order(x),,]
  lon <- lon[order(x),,]
  x <- x[order(x)]

  #Getting years where decade 0 and decade N start from
  dec0 <- results_query$decade_0_start[i]
  decN <- results_query$decade_N_start[i]

  #Converting time from integers to dates
  #Extracting date time origin
  origin <- str_extract(ncatt_get(r, "time", "units")$value, "\\d{4}-\\d{2}-\\d{2}")
  #Saving converted dates into tibble
  ds_dates <- time %>%
    as_tibble() %>%
    #Adding column ID for easy reference
    rowid_to_column("id") %>%
    #Adding month, year and decade
    mutate(time = as_date(value, origin = origin),
           year = year(time),
           month = month(time),
           #First decade classified as 1, final decade identified as 2
           dec = case_when(year <= results_query$decade_0_end[i] ~ 1,
                           T ~ 2)) %>%
    #Removing original time values
    select(-value)

  #Calculating monthly means per decade
  for(d in unique(ds_dates$dec)){
    month_mean <- list()
    for(m in 1:12){
      #Slicing dataset per month/decade
      #Identifying indices
      ind <- ds_dates %>%
        filter(dec == d & month == m) %>%
        select(id) %>%
        pull()
      #Calculating monthly means and reordering for easy plotting
      month_mean[[m]] <- apply(var[,,ind], 1:2, mean, na.rm = T)
    }
    #Creating a single file per decade
    dec_month_mean <- array(unlist(month_mean), dim = c(length(x), length(y), 12))

    #Creating netcdf
    #Adding lat and lon coordinates as indexes
    xdim <- ncdim_def("x", "degrees_east", x)
    ydim <- ncdim_def("y", "degrees_north", y)
    #Adding time dimension
    timedim <- ncdim_def("month", "month of the year", 1:12)
    #Adding monthly mean
    clim_ds <- ncvar_def(paste0(var_id, "_monthly_mean"), ncatt_get(r, var_id)$units,
                         list(xdim, ydim, timedim),
                         ncatt_get(r, var_id)$missing_value,
                         paste0("Monthly mean ",
                                ncatt_get(r, var_id)$long_name))
    #Adding lat and lon as separate dimension
    lon_ds <- ncvar_def("lon", ncatt_get(r, "lon")$units,
                         list(xdim, ydim, timedim), ncatt_get(r, "lon")$missing_value,
                         ncatt_get(r, "lon")$long_name)
    lat_ds <- ncvar_def("lat", ncatt_get(r, "lat")$units,
                         list(xdim, ydim, timedim), ncatt_get(r, "lat")$missing_value,
                         ncatt_get(r, "lat")$long_name)
    #Copying global attributes from original file
    global_atts <- ncatt_get(r, 0)


    #Naming monthly mean files per decade
    if(d == 1){
      dec_start <- dec0
    }else{dec_start <- decN}
    #Creating file name
    file_name <- paste0(results_query$base_file_name[i], ".MonthlyMean.",
                        dec_start, "-", (dec_start+9), ".nc")
    #Creating full path
    file_out <- file.path(results_query$out_full_folder[i], file_name)

    #Putting everything together in one file
    ncout <- nc_create(file_out, list(clim_ds, lon_ds, lat_ds), force_v4 = T)
    ncvar_put(ncout, clim_ds, dec_month_mean)
    ncvar_put(ncout, lon_ds, lon[,,1:12])
    ncvar_put(ncout, lat_ds, lat[,,1:12])
    for(j in 1:length(global_atts)){
      ncatt_put(ncout, 0, names(global_atts)[j], global_atts[[j]])
    }

    #Closing the file so it is saved on disk
    nc_close(ncout)
  }
  nc_close(r)
}

```


## Plotting results
We will use the results of the last item in the loop to create a simple plot with the mean monthly values for January.
```{r eval = F}
ex <- nc_open(file_out)
ex_var <- ncvar_get(ex, paste0(var_id, "_monthly_mean"))
image(ex_var[,,1])
```


