---
title: "Plotting CMIP6 ESM data in R"
author: "Denisse Fierro Arcos"
date: "2022-10-29"
output: 
  github_document:
    toc: true
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

In this notebook, we will use the `metR` and `ggplot2` libraries in `R` to load the monthly means from the CMIP6 data.


## Loading libraries

```{r libraries, results = "hide", warnings = F, message = F}
library(metR)
library(tidyverse)
```

## Loading search results from [first notebook](01_Querying_CMIP6_database.md)
We will use the search results from the first notebook to find the files we need. We will only focus on the `intpp` variable and the `ssp245` scenario.

```{r}
#Loading search results
results_query <- read_csv("../Outputs/results_merged.csv") %>%
  filter(variable_id == "intpp" & experiment_id == "ssp245") %>% 
  #We will remove any duplicated entries - based on the output file name
  mutate(dup = duplicated(out_file_name)) %>%
  filter(dup == F) %>% 
  #Add name of output files of interest
  mutate(mean_dec0 = list.files(out_full_folder, pattern = paste0("*.intpp.*Mean.", decade_0_start), full.names = T),
         mean_decN = list.files(out_full_folder, pattern = paste0("*.intpp.*Mean.", decade_N_start), full.names = T))

head(results_query, n = 2)
```



## Loading data

```{r}
var_id <- results_query$variable_id[1]
file_dec0 <- results_query$mean_dec0[1]
file_decN <- results_query$mean_decN[1]
model <- results_query$source_id[1]
months_ind <- c(1, 9)


#Function creating one raster per month of interest
raster_month <- function(df){
  df %>% 
    dplyr::select(-month) %>% 
    #Changing data frame to sf object
    st_as_sf(coords = c("x", "y"), crs = 4326) %>% 
    #Creating raster
    st_rasterize() %>% 
    #Changing attributes to dimensions
    st_redimension() %>% 
    #Renaming dimension to decades. Value 2 represents difference between first (0) and last (1) decade
    st_set_dimensions("new_dim", values = c(0,1,2), names = "decade")
  }

#Loading raster with means from first decade
raster_stack <- function(file_dec0, file_decN, var_id, months_ind){
  ReadNetCDF(file_dec0, vars = var_id) %>% 
  #Extracting months of interest
  filter(month %in% months_ind) %>% 
  #Adding a column indicating to which decade value relates to
  mutate(decade = "Decade_0") %>% 
  #Loading raster with means from last decade
  bind_rows(ReadNetCDF(file_decN, vars = var_id) %>% 
              #Extracting months of interest
              filter(month %in% months_ind) %>% 
               #Adding a column indicating to which decade value relates to
              mutate(decade = "Decade_N")) %>%
  #Calculating difference between first and last decade
  pivot_wider(names_from = decade, values_from = var_id) %>%
  mutate(Difference = Decade_0 - Decade_N,
         x = case_when(x >180 ~ x-360,
                       T~x)) %>%
  group_by(month)
  }

test <- results_query %>% 
  dplyr::select(variable_id, mean_dec0, mean_decN, source_id) %>% 
  as_tibble() %>% 
  mutate(df = NA, raster = NA, figure = NA)

for(i in 1:nrow(test)){
  test$df[i] <- list(raster_stack(test$mean_dec0[i], test$mean_decN[i], test$variable_id[i], months_ind))
  test$raster[i] <- list(group_split(test$df[i][[1]]) %>% purrr::map(raster_month))
  # n <- month.abb[group_keys(test$df[1][[1]]) %>% pull()]
  }
           
# library(patchwork)
# 
# #Applying rastering function
# r_month <- test$df[1] %>% 
#   group_split() %>% 
#   purrr::map(raster_month)
# 
# wgs84_map <- ggplot()+
#   geom_stars(data = test$raster[10][[1]])+
#   facet_grid(~decade)+
#   theme_bw()
# 
# wgs84_map2 <- ggplot()+
#   geom_stars(data = test$raster[3][[1]])+
#   facet_grid(~decade)+
#   theme_bw()
# 
# test0 <- st_transform(r_month[[1]] %>% filter(y <= 84), crs = st_crs(3395))
# 
# mercator_map <- ggplot()+
#   geom_stars(data = test0)+
#   facet_grid(~decade)+
#   theme_bw()
# 
# wgs84_map/wgs84_map2

```

