---
title: "Plotting CMIP6 ESM data in R"
author: "Denisse Fierro Arcos"
date: "2022-10-29"
output: 
  github_document:
    toc: true
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

In this notebook, we will use the `ncdf4` and `lattices` libraries in `R` to plot monthly decadal means calculated from the CMIP6 data in our previous notebooks for [`R`]("03a_Calculating_monthly_means_in_R.md") or [`Python`]("03b_Calculating_monthly_means_in_Python.md").


## Loading libraries

```{r libraries, results = "hide", warnings = F, message = F}
library(ncdf4)
library(lattice)
library(latticeExtra)
library(tidyverse)
```

## Loading search results from [first notebook](01_Querying_CMIP6_database.md)
We will use the search results from the first notebook to find the files we need. We will only focus on the `intpp` variable and the `ssp245` scenario.

```{r}
#Loading search results
results_query <- read_csv("../Outputs/results_merged.csv") %>%
  filter(variable_id == "intpp" & experiment_id == "ssp245") %>% 
  #We will remove any duplicated entries - based on the output file name
  mutate(dup = duplicated(out_file_name)) %>%
  filter(dup == F) %>% 
  #Add name of output files of interest
  mutate(mean_dec0 = list.files(out_full_folder, pattern = paste0("*.intpp.*Mean.", decade_0_start), full.names = T),
         mean_decN = list.files(out_full_folder, pattern = paste0("*.intpp.*Mean.", decade_N_start), full.names = T))

head(results_query, n = 2)
```

## Loading data
We will create a new data frame from the search results above to store our rasters and plots. We will also add a few empty columns to store our results.

```{r}
monthly_dec_means <- results_query %>% 
  dplyr::select(variable_id, mean_dec0, mean_decN, source_id) %>% 
  as_tibble() %>% 
  #The nc column will contain the rasters for each month of interest per decade (dec0 and decN), as well as the difference
  mutate(#Plots will be saved in these columns
         fig_dec0 = NA, fig_decN = NA, fig_diff = NA)
```

We will only be plotting the means for two months, for which we will include its index under the `months_ind` variable. We will be plotting January and September.

```{r}
months_ind <- c(1, 9)
```

We will define a few functions to help us speed up the process of loading and plotting data.
```{r}
#Loading nc function
loading_nc_means <- function(file_path, variable, months){
  #Loading nc file
  nc <- nc_open(file_path)
  #Extracting variables of interest
  var <- ncvar_get(nc, variable)
  #Extracting coordinates
  lon <- ncvar_get(nc, "x")
  lat <- ncvar_get(nc, "y")
  #Transforming longitude from 0-360 to -/+180
  lon_180 <- (lon[lon > 180])-360
  lon_180 <- c(lon_180, lon[lon <= 180])
  #Getting indexes from reordered longitude to reorder data
  ind <- c(which(lon>180), which(lon<=180))
  #Reordering data and extracting months of interest
  var <- var[ind,,months]
  #Putting all outputs in one list
  out <- list(raster = var,
              lon = lon_180,
              lat = lat)
  nc_close(nc)
  return(out)
}
```



We will now go through each row of our search results, load the correct `netcdf` file, calculate the difference, and plot the results.

```{r}
#Changing month indexes to month abbreviations.
#Looping through each row in our results
for(i in 1:nrow(monthly_dec_means)){
  #Loading rasters with data for both decades
  nc_dec0 <- loading_nc_means(monthly_dec_means$mean_dec0[i], 
                              monthly_dec_means$variable_id[i],
                              months_ind)
  nc_decN <- loading_nc_means(monthly_dec_means$mean_decN[i],
                              monthly_dec_means$variable_id[i], 
                              months_ind)
  #Calculate differences per month/decade
  diff <- nc_dec0$raster-nc_decN$raster
  #Extract coordinates for plotting
  lon <- nc_dec0$lon
  lat <- as.vector(nc_dec0$lat)
  #Create grid for plotting
  grid <- expand.grid(lon = lon, lat = lat)
  #Create color bar with 9 groups
  cutpts <- seq(0, max(max(nc_dec0$raster, na.rm = T), max(nc_decN$raster, na.rm = T)), length = 9)
  cutpts_diff <- c(seq(min(diff, na.rm = T), -0.1e-8, length = 6),
                   seq(0.1e-8, max(diff, na.rm = T), length = 5))
  #Create plots
  figs_dec0 <- list();  figs_decN <- list(); figs_diff <- list()
  for(m in 1:length(months_ind)){
    title <- paste0(monthly_dec_means$source_id[i], " - ", month.abb[m])
    figs_dec0[[m]] <- levelplot(nc_dec0$raster[,,m] ~ lon*lat, data = grid, pretty = T, main = monthly_dec_means$source_id[i],
                                at = cutpts, cuts = 9, col.regions = RColorBrewer::brewer.pal(9, "PuBuGn"))
    figs_decN[[m]] <- levelplot(nc_decN$raster[,,m] ~ lon*lat, data = grid, pretty = T, main = monthly_dec_means$source_id[i],
                                at = cutpts, cuts = 9, col.regions = RColorBrewer::brewer.pal(9, "PuBuGn"))
    figs_diff[[m]] <- levelplot(diff[,,m] ~ lon*lat, data = grid,  pretty = T, main = monthly_dec_means$source_id[i],
                                at = cutpts_diff, cuts = 11, col.regions = RColorBrewer::brewer.pal(11, "RdYlBu"))
  }
  
  
  
  monthly_dec_means$fig_dec0[i] <- list(dec0 = levelplot(nc_dec0$raster[,,m] ~ lon*lat, data = grid, 
                                                         at = cutpts, cuts = 9, pretty = T,
                                                         col.regions = RColorBrewer::brewer.pal(9, "PuBuGn")),
                                        decN = levelplot(nc_decN$raster[,,m] ~ lon*lat, data = grid, 
                                                         at = cutpts, cuts = 9, 
                                                         col.regions = RColorBrewer::brewer.pal(9, "PuBuGn")),
                                        diff = levelplot(diff[,,m] ~ lon*lat, data = grid, 
                                                         at = cutpts, cuts = 9, 
                                                         col.regions = RColorBrewer::brewer.pal(9, "PuBuGn")))
  b
  x <- nc_dec0$raster-nc_decN$raster
  monthly_dec_means$nc_dec0[1][[1]]$raster[,,1]
  
  monthly_dec_means$nc_dec0[1][[1]]$lat
  #Save data frame with combined datasets
  monthly_dec_means$nc[i]
    df[i] <- list(raster_stack(test$mean_dec0[i], 
                                  test$mean_decN[i], 
                                  test$variable_id[i], 
                                  months_ind))
  
  
 
#Calculate differences
diff <- var[,,1]-var[,,2]
image(lon_180, lat, diff, col = RColorBrewer::brewer.pal(11, "RdYlBu"));
contour(lon_180, lat, diff, add = T, lty = 0.9);
title(main = "test_model", sub = "test", font.main = 4);box()
```
```{r}
#More plots
library(latticeExtra)
grid <- expand.grid(lon_180 = lon_180, lat = lat)
cutpts <- seq(min(diff, na.rm = T), max(diff, na.rm = T), length = 11)
b <- lattice::levelplot(diff ~ lon_180*lat, data = grid, at = cutpts, cuts = 11, 
                   col.regions=rev(RColorBrewer::brewer.pal(11, "RdYlBu")))
b
c(b, b, layout = c(2, 2))
as.layer(b)

nc_close(nc)
```

