---
title: "Plotting CMIP6 ESM data in R"
author: "Denisse Fierro Arcos"
date: "2022-10-29"
output: 
  github_document:
    toc: true
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

In this notebook, we will use the `metR` and `ggplot2` libraries in `R` to load the monthly means from the CMIP6 data.


## Loading libraries

```{r libraries, results = "hide", warnings = F, message = F}
library(metR)
library(tidyverse)
```

## Loading search results from [first notebook](01_Querying_CMIP6_database.md)
We will use the search results from the first notebook to find the files we need. We will only focus on the `intpp` variable and the `ssp245` scenario.

```{r}
#Loading search results
results_query <- read_csv("../Outputs/results_merged.csv") %>%
  filter(variable_id == "intpp" & experiment_id == "ssp245") %>% 
  #We will remove any duplicated entries - based on the output file name
  mutate(dup = duplicated(out_file_name)) %>%
  filter(dup == F) %>% 
  #Add name of output files of interest
  mutate(mean_dec0 = list.files(out_full_folder, pattern = paste0("*.intpp.*Mean.", decade_0_start), full.names = T),
         mean_decN = list.files(out_full_folder, pattern = paste0("*.intpp.*Mean.", decade_N_start), full.names = T))

head(results_query, n = 2)
```



## Loading data

```{r}
months_ind <- c(1, 9)

#Function creating one raster per month of interest
raster_month <- function(df){
  df %>% 
    dplyr::select(-month) %>% 
    #Changing data frame to sf object
    st_as_sf(coords = c("x", "y"), crs = 4326) %>% 
    #Creating raster
    st_rasterize() %>% 
    #Changing attributes to dimensions
    st_redimension() %>% 
    #Renaming dimension to decades. Value 2 represents difference between first (0) and last (1) decade
    st_set_dimensions("new_dim", values = c(0,1,2), names = "decade")
  }

#Loading raster with means from first decade
raster_stack <- function(file_dec0, file_decN, var_id, months_ind){
  ReadNetCDF(file_dec0, vars = var_id) %>% 
  #Extracting months of interest
  filter(month %in% months_ind) %>% 
  #Adding a column indicating to which decade value relates to
  mutate(decade = "Decade_0") %>% 
  #Loading raster with means from last decade
  bind_rows(ReadNetCDF(file_decN, vars = var_id) %>% 
              #Extracting months of interest
              filter(month %in% months_ind) %>% 
               #Adding a column indicating to which decade value relates to
              mutate(decade = "Decade_N")) %>%
  #Calculating difference between first and last decade
  pivot_wider(names_from = decade, values_from = var_id) %>%
  mutate(Difference = Decade_0 - Decade_N,
         x = case_when(x >180 ~ x-360,
                       T~x)) %>%
  group_by(month)
  }

test <- results_query %>% 
  dplyr::select(variable_id, mean_dec0, mean_decN, source_id) %>% 
  as_tibble() %>% 
  mutate(df = NA, raster = NA, figure = NA)

#Facet labels
label_names <- c(`0` = "First decade",
                   `1` = "Last decade",
                   `2` = "Difference")

for(i in 1:nrow(test)){
  # test$df[i] <- list(raster_stack(test$mean_dec0[i], test$mean_decN[i], test$variable_id[i], months_ind))
  # test$raster[i] <- list(group_split(test$df[i][[1]]) %>% purrr::map(raster_month) %>% setNames(month.abb[months_ind]))
  figs <- list(jan_fig = ggplot()+
                 geom_stars(data = test$raster[i][[1]]$Jan)+
                 facet_grid(~decade, labeller = as_labeller(label_names))+
                 scale_fill_viridis_c(guide = guide_colorbar(title = test$variable_id[i],
                                             title.position = "top",
                                             title.hjust = 0.5))+
                 theme_bw()+
                 theme(legend.position = "bottom", 
                       legend.key.width = unit(3, "cm"),
                       axis.title = element_blank()),
    sep_fig = ggplot()+
      geom_stars(data = test$raster[i][[1]]$Sep)+
      facet_grid(~decade, labeller = as_labeller(label_names))+
                 scale_fill_viridis_c(guide = guide_colorbar(title = test$variable_id[i],
                                             title.position = "top",
                                             title.hjust = 0.5))+
                 theme_bw()+
                 theme(legend.position = "bottom", 
                       legend.key.width = unit(3, "cm"),
                       axis.title = element_blank()))
  test$figure[i] <- list(figs)
}

```

