---
title: "Plotting CMIP6 ESM data in R"
author: "Denisse Fierro Arcos"
date: "2022-10-29"
output: 
  github_document:
    toc: true
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

In this notebook, we will use the `ncdf4` and `lattices` libraries in `R` to plot monthly decadal means calculated from the CMIP6 data in our previous notebooks for [`R`]("03a_Calculating_monthly_means_in_R.md") or [`Python`]("03b_Calculating_monthly_means_in_Python.md").


## Loading libraries

```{r libraries, results = "hide", warnings = F, message = F}
library(tidyverse)
library(reticulate)
```

## Loading search results from [first notebook](01_Querying_CMIP6_database.md)
We will use the search results from the first notebook to find the files we need. We will only focus on the `intpp` variable and the `ssp245` scenario. This first part of the script is done in `R` because it offers an easy way to manipulate tabular data.

```{r}
#Loading search results
results_query <- read_csv("../Outputs/results_merged.csv") %>%
  filter(variable_id == "intpp" & experiment_id == "ssp245") %>% 
  #We will remove any duplicated entries - based on the output file name
  distinct(out_file_name, .keep_all = T) %>% 
  mutate(mean_dec0 = NA,
         mean_decN = NA)
```

```{r}
for(i in 1:nrow(results_query)){
  results_query$mean_dec0[i] = list.files(results_query$out_full_folder[i],
                                          pattern = paste0("*.intpp.*Mean.", results_query$decade_0_start[i]), 
                                          full.names = T)
  results_query$mean_decN[i] = list.files(results_query$out_full_folder[i],
                                          pattern = paste0("*.intpp.*Mean.", results_query$decade_N_start[i]), 
                                          full.names = T)
  
}

#We can check results
head(results_query, n = 2)
```
## Loading data using Python
We will use the `reticulate` package to load `Python` into our environment.

```{r libraries, results = "hide", warnings = F, message = F}
use_condaenv("CMIP6_data")
```
## Loading relevant Python libraries
We will now load relevant libraries that will allow us to load `netcdf` files and prepare plots easily.

```{python}
import xarray as xr
import os
from glob import glob
import calendar
import pandas as pd
import numpy as np
from itertools import cycle

#Packages for plotting
import matplotlib.pyplot as plt
import matplotlib.ticker as mticker
import matplotlib.dates as mdates
import matplotlib.colors as mcolors
from matplotlib.ticker import AutoMinorLocator, MultipleLocator
from cartopy.mpl.gridliner import LONGITUDE_FORMATTER, LATITUDE_FORMATTER
import cmocean as cm
import cartopy.crs as ccrs
import cartopy.feature as cft
import matplotlib.gridspec as gridspec
import matplotlib.cm as mcm
```
We will load the query results that we processed at the beginning of this script. This will help us identify the files that are relevant to our work. We will also define a new variable called `months_interest` where we will include the months that we will be plotting in our figures. In this case, we will plot the months of January and September.

```{python}
#We will only keep a handful of columns that are relevant to our plots.
results_query = r.results_query[['variable_id', 'mean_dec0', 'mean_decN', 'source_id']]

#We define the months that are interested in plotting.
months_interest = ['January', 'September']

```

The datasets contain a `months` dimension which indicates to which month the data is linked to. But, it uses numbers to represent the month. We will update this names are used instead.

```{python}
data = {}
for i in range(len(results_query)):
  #Loading decadal means and ensuring only positive values are kept
  dec0 = xr.open_dataset(results_query.mean_dec0[i])
  dec0 = dec0.where(dec0 >= 0)
  decN = xr.open_dataset(results_query.mean_decN[i])
  decN = decN.where(decN >= 0)
  #Calculating differences between decades
  diff = dec0-decN
  #Stacking everything together - Decade dimension added to identify data easily
  stack = xr.concat([dec0.intpp.expand_dims({'decade': ['dec0']}), 
  decN.intpp.expand_dims({'decade': ['decN']}), 
  diff.intpp.expand_dims({'decade': ['diff']})], dim = 'decade')
  #Changing months values from digits to names
  stack['month'] = [calendar.month_name[m] for m in stack.month.values]
  #Correcting longitude from 0 to 360 to -180 to +180
  #lon = stack.x.values
  #lon[lon > 180] = lon[lon > 180]-360
  #stack['x'] = sorted(lon)
  model = results_query.source_id[i]
  #Keeping months of interest only
  data[model] = stack.loc[:, months_interest]

```

#Finding range of values in data to create colorbar for plots

```{python}
min_intpp = []
max_intpp = []
min_diff = []
max_diff = []
for mod in data:
  min_intpp.append(float(data[mod][0:2].min().values))
  min_diff.append(float(data[mod][2].min().values))
  max_intpp.append(float(data[mod][0:2].max().values))
  max_diff.append(float(data[mod][2].max().values))
  
min_intpp = min(min_intpp)
max_intpp = max(max_intpp)
min_diff = min(min_diff)
max_diff = max(max_diff)
```

Finally, we will plot the subsetted data.
```{python}
#Selecting month to be plotted
month_plot = months_interest[0]

#Initialising figure
fig = plt.figure(figsize = (10, 30))

#Create colormap for differences
#Ensure colormap diverges at zero regardless of max and min values
divnorm = mcolors.TwoSlopeNorm(vmin = min_diff*1.1, vcenter = 0, \
vmax = max_diff*1.1)
#Create new diverging colormap from existing colormaps to make it
#easy to distinguish positive and negative values
colors1 = plt.cm.magma(np.linspace(0.25, 1, 128))
colors2 = cm.cm.ice_r(np.linspace(0.1, 0.75, 128))
colors = np.vstack((colors1, colors2))
mymap = mcolors.LinearSegmentedColormap.from_list('my_colormap', colors)
#Set NA values to show as white
mymap.set_bad('white')

#Set font family and font size for the entire graph
plt.rcParams['font.family'] = 'serif'
plt.rcParams['font.size'] = 8

#Create variable containing the Antarctic continent - 50m is medium scale
land_50m = cft.NaturalEarthFeature('physical', 'land', '50m', edgecolor = 'black', 
                                   facecolor = 'gray', 
                                   linewidth = 0.5)
                                   
#Defining groups for plots
decadeName = {'dec0': 'First decade', 
              'decN': 'Last decade', 
              'difference': 'Difference (first - last)'}

#Getting names of models to be plotted
models = [mod for mod in data.keys()]

#Merge seasons and datasets
counter = zip(cycle(decadeName), np.repeat(models, len(decadeName)))

########
#Create composite figure using a grid
#Height ratios - all models the same, last row 1/10 (for colorbar)
height_rat = [1]*len(models)
height_rat.append(0.1)
gs = gridspec.GridSpec(nrows = len(models)+1,
                       ncols = len(decadeName),
                       height_ratios = height_rat)
#Change the width between each plot
gs.update(wspace = 0.05, hspace = 0.05)

#Initialising axes for plot 
ax = {}

#Creating plot using loops
for i, coord in enumerate(counter):
  ax[i] = fig.add_subplot(gs[i], projection = ccrs.PlateCarree())
  #Initial decade data
  if coord[0] == 'dec0':
    da = data[coord[1]].sel(month = month_plot, decade = coord[0])
    p1 = da.plot.pcolormesh(ax = ax[i], cmap = cm.cm.speed,
                   add_colorbar = False)
  #Last decade
  # elif coord[0] == 'decN':
  #   da_N = data[coord[1]].sel(month = month_plot, decade = coord[0])
  #   p1 = da_N.plot.pcolormesh(ax = ax[i], cmap = cm.cm.speed,
  #                add_colorbar = False)
  # #Differences
  # else:
  #   #Remove zeroes
  #   da_d = data[coord[1]].sel(month = month_plot, decade = coord[0])
  #   #Plot data
  #   p2 = da_d.plot(ax = ax[i], cmap = mymap, norm = divnorm,
  #   add_colorbar = False)
  # 
  # #Set title for each map
  # if i < 3 and coord[0] == 'decN':
  #   ax[i].set_title(f'{decadeName[coord[0]]} \n {coord[1]}', y = 1.01)
  # elif i < 3:
  #   ax[i].set_title(f'{decadeName[coord[0]]} \n', y = 1.01)
  # elif i > 3 and coord[0] == 'decN':
  #   ax[i].set_title(f'{coord[1]}', y = 1.01)
  # else:
  #   ax[i].set_title('')
        
  #Add land and coastlines
  ax[i].add_feature(land_50m)
  ax[i].coastlines(resolution = '50m')
    
# ax_col = fig.add_subplot(gs[-1, 0:2])
# cb = plt.colorbar(p1, cax = ax_col, orientation = 'horizontal', 
#                   shrink = 0.5,
#                   label = 'Primary Organic Carbon Production (mol m-2 s-1)')
# 
# ax_dif = fig.add_subplot(gs[-1, -1])
# cb_dif = plt.colorbar(p2, cax = ax_dif, orientation = 'horizontal', 
#                   extend = 'both', shrink = 0.5,
#                   label = 'Difference (mol m-2 s-1)')
# 
# plt.savefig('../Figures/test.png', dpi = 300, bbox_inches = 'tight', pad_inches = 0.05)
  #os.path.join(folder_out, fn), dpi = 300, 
   #         bbox_inches = 'tight', pad_inches = 0.05)
plt.show()


```



We will now go through each row of our search results, load the correct `netcdf` file, calculate the difference, and plot the results.

```{r}
#Changing month indexes to month abbreviations.
#Looping through each row in our results
month_Jan <- list()
month_Sep <- list()
for(i in 1:nrow(monthly_dec_means)){
  dec0 <- loading_nc_means(monthly_dec_means$mean_dec0[i], 
                              monthly_dec_means$variable_id[i],
                              months_ind)
  decN <- loading_nc_means(monthly_dec_means$mean_decN[i],
                              monthly_dec_means$variable_id[i],
                              months_ind)
  diff <- dec0$raster-decN$raster
  name <- monthly_dec_means$source_id[i]
  month_Jan[[name]] <- list(dec0 = dec0$raster[,,1],
                          decN = decN$raster[,,1],
                          diff = diff[,,1],
                          lon = dec0$lon,
                          lat = dec0$lat)
  month_Sep[[name]] <- list(dec0 = dec0$raster[,,2],
                          decN = decN$raster[,,2],
                          diff = diff[,,2],
                          lon = dec0$lon,
                          lat = dec0$lat)

  #Loading rasters with data for both decades
  # dec_0[[i]] <- list(name = monthly_dec_means$source_id[i],
  #                    data = list(loading_nc_means(monthly_dec_means$mean_dec0[i], 
  #                             monthly_dec_means$variable_id[i],
  #                             months_ind)))
  # dec_N[[i]] <- list(name = monthly_dec_means$source_id[i],
  #                    data = loading_nc_means(monthly_dec_means$mean_decN[i],
  #                             monthly_dec_means$variable_id[i],
  #                             months_ind))
  # diff[[i]] <- dec_0[[i]]$data$raster-dec_0[[i]]$data$raster
  # 
  # <- loading_nc_means(monthly_dec_means$mean_dec0[i], 
  #                             monthly_dec_means$variable_id[i],
  #                             months_ind)
  # dec_N <- loading_nc_means(monthly_dec_means$mean_decN[i],
  #                             monthly_dec_means$variable_id[i],
  #                             months_ind)
  # monthly_dec_means$dec0_raster[i] <- list(raster = dec_0$raster)
  # monthly_dec_means$decN_raster[i] <- list(raster =dec_N$raster)
  # monthly_dec_means$coordinates[i] <- list(lon = dec_0$lon, 
  #                                          lat = dec_0$lat)
  # monthly_dec_means$coordinates[i] <- list(loading_nc_means(monthly_dec_means$mean_decN[i],
  #                             monthly_dec_means$variable_id[i],
  #                             months_ind, "coordinates"))
}

  
  #Calculate differences per month/decade
  # diff <- nc_dec0$raster-nc_decN$raster
  # #Extract coordinates for plotting
  # lon <- nc_dec0$lon
  # lat <- as.vector(nc_dec0$lat)
  # #Create grid for plotting
  # grid <- expand.grid(lon = lon, lat = lat)
  # monthly_dec_means$dec0_Jan[i] = nc_dec0$raster[,,"jan"]
  # monthly_dec_means$dec0_Sep[i] = nc_dec0$raster[,,"sep"]

  
#Create color bar with 9 groups
  cutpts <- seq(0, max(max(nc_dec0$raster, na.rm = T), max(nc_decN$raster, na.rm = T)), length = 9)
  cutpts_diff <- c(seq(min(diff, na.rm = T), -0.1e-8, length = 6),
                   seq(0.1e-8, max(diff, na.rm = T), length = 5))
  #Create plots
  figs_dec0 <- list();  figs_decN <- list(); figs_diff <- list()
  for(m in 1:length(months_ind)){
    title <- paste0(monthly_dec_means$source_id[i], " - Decadal mean for", month.name[m] )
    #Changing NA values to -9999 to color land in plots
    nc_dec0$raster[,,m][is.na(nc_dec0$raster[,,m])] <- -9999
    nc_decN$raster[,,m][is.na(nc_decN$raster[,,m])] <- -9999
    diff[,,m][is.na(diff[,,m])] <- -9999
    #Plotting results
    figs_dec0[[m]] <- levelplot(nc_dec0$raster[,,m] ~ lon*lat, data = grid, pretty = T, main = title, at = c(-Inf, cutpts), 
                                cuts = 10, col.regions = c("#dcd0c4", RColorBrewer::brewer.pal(9, "PuBuGn")),
                                contour = T, col = "#a6a6a6")
    figs_decN[[m]] <- levelplot(nc_decN$raster[,,m] ~ lon*lat, data = grid, pretty = T, main = title, at = c(-Inf, cutpts), 
                                cuts = 10, col.regions = c("#dcd0c4", RColorBrewer::brewer.pal(9, "PuBuGn")),
                                contour = T, col = "#a6a6a6")
    figs_diff[[m]] <- levelplot(diff[,,m] ~ lon*lat, data = grid,  pretty = T, main = title, at = c(-Inf, cutpts_diff), 
                                cuts = 12, col.regions = c("#dcd0c4",RColorBrewer::brewer.pal(11, "RdYlBu")),
                                contour = T, col = "#a6a6a6")
  #Update names of figures to month abbreviation
  names(figs_dec0) <- month.abb[months_ind]
  #names(figs_decN) <- month.abb[months_ind]
  #names(figs_diff) <- month.abb[months_ind]
  #Saving figures in tibble
  monthly_dec_means$fig_dec0[i] <- list(figs_dec0)
  monthly_dec_means$fig_decN[i] <- list(figs_decN)
  monthly_dec_means$fig_diff[i] <- list(figs_diff)  
  }
  
  
  # monthly_dec_means$fig_dec0[i] <- list(dec0 = levelplot(nc_dec0$raster[,,m] ~ lon*lat, data = grid, 
  #                                                        at = cutpts, cuts = 9, pretty = T,
  #                                                        col.regions = RColorBrewer::brewer.pal(9, "PuBuGn")),
  #                                       decN = levelplot(nc_decN$raster[,,m] ~ lon*lat, data = grid, 
  #                                                        at = cutpts, cuts = 9, 
  #                                                        col.regions = RColorBrewer::brewer.pal(9, "PuBuGn")),
  #                                       diff = levelplot(diff[,,m] ~ lon*lat, data = grid, 
  #                                                        at = cutpts, cuts = 9, 
  #                                                        col.regions = RColorBrewer::brewer.pal(9, "PuBuGn")))
  # b
  # x <- nc_dec0$raster-nc_decN$raster
  # monthly_dec_means$nc_dec0[1][[1]]$raster[,,1]
  # 
  # monthly_dec_means$nc_dec0[1][[1]]$lat
  # #Save data frame with combined datasets
  # monthly_dec_means$nc[i]
  #   df[i] <- list(raster_stack(test$mean_dec0[i], 
  #                                 test$mean_decN[i], 
  #                                 test$variable_id[i], 
  #                                 months_ind))
  # 
  
 
#Calculate differences
diff <- var[,,1]-var[,,2]
image(lon_180, lat, diff, col = RColorBrewer::brewer.pal(11, "RdYlBu"));
contour(lon_180, lat, diff, add = T, lty = 0.9);
title(main = "test_model", sub = "test", font.main = 4);box()
```
```{r}
#More plots
library(latticeExtra)
grid <- expand.grid(lon_180 = lon_180, lat = lat)
cutpts <- seq(min(diff, na.rm = T), max(diff, na.rm = T), length = 11)
b <- lattice::levelplot(diff ~ lon_180*lat, data = grid, at = cutpts, cuts = 11, 
                   col.regions=rev(RColorBrewer::brewer.pal(11, "RdYlBu")))
b
c(b, b, layout = c(2, 2))
as.layer(b)

nc_close(nc)
```

