---
title: "Accessing CMIP6 ESM data"
author: "Denisse Fierro Arcos"
date: "2022-10-12"
output: 
  github_document:
    toc: true
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
In this notebook we will access Coupled Model Intercomparison Project Phase 6 (CMIP6) data from the Earth System Grid Federation ([ESGF](https://esgf.nci.org.au/search/cmip6-nci/)) platform.  
  
We will be using some functions from the [epwshiftr](https://github.com/ideas-lab-nus/epwshiftr) package, which have been modified slightly to access ocean and sea ice outputs.
  
Once the correct datasets have been identified, a subset of the dataset will be selected and downloaded to your local machine. Here we have chosen to select the first and last decades for each dataset, that is from 1850 to 1859 and from 2005 to 2014 for the historical runs, and from 2015 to 2024 and from 2091 to 2100 for the shared socioeconomic pathways (SSPs). Future scenario runs up to 2300 have been excluded here.
  
  
## Loading libraries
The functions adapted from `epwshiftr`, which are on the accompanying `search_download_functions` script will help us connect to the Australian-based ESGF node hosted by the National Computational Infrastructure ([NCI](http://nci.org.au/)). There are several ESGF nodes available worldwide and they can be found [here](https://aims2.llnl.gov/nodes). If you wish to change the node used in the script use the `node_url` parameter of the `cmip6_index` function to provide the link to the search page for the node. The link should look like this: [https://esgf.nci.org.au/esg-search/search/?](https://esgf.nci.org.au/esg-search/search/?) and it points to a `xml` file, which is similar to a table of contents for the datasets available within the node.
  
All other libraries allow us to manipulate and save CMIP6 data.

```{r libraries, results = "hide", warnings = F, message = F}
search_CMIP6 <- source("search_download_functions.R")
library(tidyverse)
# library(purrr)
# library(metR)
library(reticulate)
```

## Querying ESGF server
We will use the `cmip6_index` function included in the accompanying `search_download_functions` script to search the Australian node of ESGF. We save the results as a table in our environment.  
  
There are thousands of CMIP6 endorsed model outputs available through ESGF, so providing as many parameters as possible in the database search is important to narrow down results. You could refer to the following websites for more information:  
- [CMIP6 documentation](https://github.com/ES-DOC/esdoc-docs/tree/master/cmip6/experiments/spreadsheet)
- [CMIP6 controlled vocabulary](https://github.com/WCRP-CMIP/CMIP6_CVs)
- [World Climate Research Programme (WCRP)](https://www.wcrp-climate.org/modelling-wgcm-mip-catalogue/modelling-wgcm-cmip6-endorsed-mips)
- [CMIP6 overview](https://pcmdi.llnl.gov/CMIP6/)
- [CMIP6 Guidance for Data Users](https://pcmdi.llnl.gov/CMIP6/Guide/dataUsers.html)
- [CMIP6 Data Request](https://clipc-services.ceda.ac.uk/dreq/mipVars.html)
- [CMIP6 Global Attributes Definitions](https://docs.google.com/document/d/1h0r8RZr_f3-8egBMMh7aqLwy3snpD6_MrDz1q8n5XUk/edit)
  
Note that parameters are compulsory, we can provide as many or as little as needed.

```{r server_query}
results_query <- cmip6_index(
  # This parameter relates to the broad experiment group a dataset belongs to. In this example we consider:
  # CMIP (past conditions) and ScenarioMIP (future scenarios, SSPs) activities
  activity = c("ScenarioMIP", "CMIP"),
  
  # Standardised name for variables of interest - See CMIP6 Data Request for more information
  variable = c("tos", "siconc", "intpp"),
  
  # This ranges from hours to years - See CMIP6 Data Request for frequency available for each variable
  frequency = "mon",
  
  # Specifying experiment names - Descriptions are available in the CMIP6 documentation
  experiment = c("ssp126", "ssp245", "ssp585", "historical"),
  
  # Specifying model names - A list of all models can be found in the CMIP6 controlled vocabulary link
  source = "ACCESS-ESM1-5",
  
  # Specifying variant, which refers to the model run 
  variant = "r1i1p1f1",
  
  # Specifying time frame of interest
  years = NULL,
  
  # Save to data dictionary
  save = TRUE
  
  # This is how we specify the node that we want to use. The default node is NCI (Australia).
  #node_url = NULL
)

#We will only show the first three results of our search.
head(results_query, n = 3)
```

### Reviewing and tidying up search results
We will add a few extra columns into our results data frame that will help us automate the extraction and storage of the data we need. We will also ensure that any datasets beyond 2100 are removed from our results.

For this example, we are interested in extracting data for the first and last decade of each experiment.

But first, we will provide the file path to the folder where we will save our data.

```{r}
out_folder <- "/perm_storage/home/data/CMIP6_data"
```

Now we will add the columns we need to automate process into the search results data frame.

```{r results_review}
results_query <- results_query %>%
  #Removing any datasets ending after 2100
  filter(year(datetime_end) <= 2100) %>% 
  #Defining the first and last decade of each experiment
  mutate(decade_0_start = year(datetime_start),
         decade_0_end = decade_0_start+9L,
         decade_N_start = year(datetime_end)-9L,
         decade_N_end = year(datetime_end),
         #Defining the name of the model outputs that we will save locally
         out_file_name = paste0(str_split(dataset_id, pattern = "\\|", 
                                          simplify = T)[1,1],
                                ".", decade_0_start, "-", decade_0_end,
                                "_", decade_N_start, "-", decade_N_end,
                                ".nc"),
         #Defining the correct file path for local copies
         out_full_folder = file.path(out_folder, source_id, experiment_id),
         #Defining full file path for subset data
         out_path = file.path(out_full_folder, out_file_name))

head(results_query, n = 3)
```

## Accessing and downloading subsets of CMIP6 data
We will now switch to `Python` to make use of the `xmip` library that allow us to standardise variable names for CMIP6 data.

### Switching to `Python`
The `reticulate` package in `R` allow us to use `Python` within an `R` script. We will be using a `conda environment` which in simple terms is a directory that contains all the `Pyhton` libraries and the dependencies needed to run this script. The `README` file in this repository has instructions on how to create this environment.

```{r}
#Activating the conda environment containing relevant Python libraries
use_condaenv("CMIP6_data")
```

### Loading `Python` libraries
```{python}
#Easy access to data hosted online
import requests

#Loading and manipulating netcdf files
from netCDF4 import Dataset
import xarray as xr

#Standardisation of CMIP6 data for easy data post-processing
from xmip.preprocessing import rename_cmip6, replace_x_y_nominal_lat_lon

#Dealing with file paths
import os
```

### Transforming CMIP6 search results into a `Python` variable
```{python}
CMIP6_query = r.results_query
CMIP6_query.head(n = 2)
```

### Accessing and saving CMIP6 data
We will now load the CMIP6 data into a temporary file in memory, so we can subset it, standardise its variables names and save a local copy. We will loop through each item in the search results.
```{python}
#Starting loop
for i in CMIP6_query.index:
  #Getting key information to load CMIP6 data locally from query results
  #URL address (include "#mode=bytes" at the end of the URL if not working)
  url = CMIP6_query['file_url'][i]
  #Variable name
  var_id = CMIP6_query['variable_id'][i]
  #Start decades
  dec_0 = CMIP6_query['decade_0_start'][i]
  dec_N = CMIP6_query['decade_N_start'][i]
  #File paths
  out_folder = CMIP6_query['out_full_folder'][i]
  out_file = CMIP6_query['out_path'][i]
  
  #Loading data to memory in a temporary file
  link = requests.get(url).content
  remote = Dataset(var_id, memory = link)
  
  #Loading as xarray dataset
  ds = xr.open_dataset(xr.backends.NetCDF4DataStore(remote))
  
  #Standardising CMIP6 data
  ds = rename_cmip6(ds)
  ds = replace_x_y_nominal_lat_lon(ds)
  
  #Subsetting data - First and last decade
  ds_sub = xr.concat([ds.sel(time = slice(str(dec_0), str(dec_0+9))),\
  ds.sel(time = slice(str(dec_N), str(dec_N+9)))], dim = 'time')
  
  #Ensure folder path exists
  os.makedirs(out_folder, exist_ok = True)
  
  #Saving subsetted dataset
  ds.to_netcdf(out_file)
  
```









## Downloading datasets
Once we are happy with our results, we are ready to download the datasets. We will keep them as temporary files only because we only need to store a subset of the original data on file.

```{r download_temporary}

# results_query %>% 
#   map(download.file(url = .$file_url, destfile = .$temp_file))

download.file(url = results_query$file_url[1], destfile = results_query$temp_file[1])
```

## Checking contents of netcdf
```{r}
GlanceNetCDF(idx$out_name[2])
```

## Loading dataset with `metR`

```{r}
ds <- ReadNetCDF(idx$out_name[2], vars = idx$variable_id[2])

#Subsetting dataset: First and last available decades
dec0 <- ds %>% filter(lubridate::year(time) <= lubridate::year(idx$datetime_start[2])+9)
decN <- ds %>% filter(lubridate::year(time) >= lubridate::year(idx$datetime_end[2])-9)
```

## Loading dataset with `stars`
```{r}
library(stars)
dss <- read_stars(results_query$temp_file[1])
dss_dim <- st_dimensions(dss)
dss_0 <- dss[, , , seq(1, length(dss_dim$time$values))[lubridate::year(dss_dim$time$values) <= 1859]]

write_stars(dss_0, "../CMIP6_data/test.nc")
```


```{python}
import xarray as xr
from netCDF4 import Dataset
import requests
from xmip.preprocessing import rename_cmip6

url = 'http://esgf.nci.org.au/thredds/fileServer/master/CMIP6/CMIP/CSIRO/ACCESS-ESM1-5/historical/r1i1p1f1/Omon/intpp/gn/v20191115/intpp_Omon_ACCESS-ESM1-5_historical_r1i1p1f1_gn_185001-201412.nc#mode=bytes'
data = requests.get(url).content

link = Dataset('intpp', memory=data)

xr.open_dataset(data, engine = 'netcdf4')
link
[print(dim) for dim in link.dimensions.values()]
[print(dim) for dim in link.variables.values()]
x = xr.open_dataset(xr.backends.NetCDF4DataStore(link))
x

```

