---
title: "Accessing CMIP6 ESM data"
author: "Denisse Fierro Arcos"
date: "2022-10-12"
output: 
  github_document:
    toc: true
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
In this notebook we will access Coupled Model Intercomparison Project Phase 6 (CMIP6) data from the Earth System Grid Federation ([ESGF](https://esgf.nci.org.au/search/cmip6-nci/)) platform.  
  
We will be using some functions from the [epwshiftr](https://github.com/ideas-lab-nus/epwshiftr) package, which have been modified slightly to access ocean and sea ice outputs.

## Loading libraries

```{r libraries, results = "hide", warnings = F, message = F}
search_CMIP6 <- source("search_download_functions.R")
library(tidyverse)
library(metR)
```

## Examples

```{r}
idx <- cmip6_index(
  # Including CMIP (past conditions) and ScenarioMIP (projections) activities
  activity = c("ScenarioMIP", "CMIP"),
  
  # SST, Primary Organic Carbon Production, SIC
  variable = c("tos", "siconc", "intpp"),
  
  # specify report frequent
  frequency = "mon",
  
  # specifying experiment names
  experiment = c("ssp126", "ssp245", "ssp585", "historical"),
  
  # specify GCM name
  source = "ACCESS-ESM1-5",
  
  # specify variant,
  variant = "r1i1p1f1",
  
  # specify years of interest
  years = NULL,
  
  # save to data dictionary
  save = TRUE
)

idx
```
### Preparing search results to download data

```{r}
idx <- idx %>% 
  #Adding a filepath for files to be downloaded as temporary files
  mutate(out_name = file.path(tempdir(), 
                              str_split(file_id, pattern = "\\|", simplify = T)[1,1])) %>% 
  #Removing any datasets ending after 2100
  filter(lubridate::year(datetime_end) <= 2100)

idx
```

## Downloading files

```{r}
download.file(idx$file_url[1], idx$out_name[1])
```

## Checking contents of netcdf
```{r}
GlanceNetCDF(idx$out_name[1])
```

## Loading dataset with `metR`

```{r}
ds <- ReadNetCDF(idx$out_name[1], vars = idx$variable_id[1])

#Subsetting dataset: First and last available decades
dec0 <- ds %>% filter(lubridate::year(time) <= lubridate::year(idx$datetime_start[1])+9)
decN <- ds %>% filter(lubridate::year(time) >= lubridate::year(idx$datetime_end[1])-9)
```

## Loading dataset with `stars`
```{r}
library(stars)
dss <- read_stars(idx$out_name[1])
dss_dim <- st_dimensions(dss)
dss_0 <- dss[, , , seq(1, length(dss_dim$time$values))[lubridate::year(dss_dim$time$values) <= 1859]]

write_stars(dss_0, "../CMIP6_data/test.nc")
```






